Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-24 paper.tex
Improved section on business. Other parties still need to finish their paragraphs. References:
iBabel - http://homepage.mac.com/swain/Sites/Macinchem/page65/ibabel3.html
ChemSpotlight - http://chemspotlight.openmolecules.net/
ChemSpider/ChemDoodle - http://www.chemspider.com/blog/ichemlabs-and-rsc-chemspider-announce-partnership.html

Although the software developed by the Blue Obelisk is Open Source on
principle, this does not rule out the establishment of a business
based on open source chemistry software. Rather than supporting itself
through sales, such a business generally relies on providing support,
customisation and services around the software. The interaction
between an open source project and a company is important, and works
best where there is some form of `giving back' to the project (such as
developer time, code contributions, or financial assistance with server costs).

>>>>

Open Source provides a unique opportunity for commercial organizations to work with the community. As opposed to traditional business models which focus on maximizing revenue through monetization of all aspects of software, working with the community allows a business to maximize profits through greater community adoption and popularity. This is an ideal solution for some commercial organizations because only they can afford the significant investment in funding and time that is required to develop a successful open source project. Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry. A goal of the Blue Obelisk is to reduce these challenges and bring commercial organizations and the community together to benefit science.

Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license. With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources. Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development. Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it. These benefits allow a company to create a successful product with significantly reduced cost and effort.

Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded. In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass. This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities. On the other hand, commercial organizations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption. In return, that commercial organization gains recognition from that product and is the first choice for commercial support, consulting and custom development. This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.

In the scientific industry, it is incredibly difficult to start successful open projects. This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment. Because of this, there is an increased barrier to starting open projects in the scientific industry. The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects. Commercial organizations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.

For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing. The license is chosen to best ensure the success of the project. iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception. iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilize the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content. Other open source developers can benefit from this library freely and expect continued development, while commercial organizations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market. Of course, no one will need to waste time rewriting functionality that is already provided with quality. Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
